Increasing prevalence of plant-fungal symbiosis across two centuries of environmental change

### 2025-10-06 Manuscript/Reviewer_response_endo_herbarium_R3.tex

\comment{Results are discussed with a very host-centric perspective. In particular, I didn't find consideration of the effect of stress on transmissibility. A relevant paper on the subject is: Gundel PE, Garibaldi LA, Martínez-Ghersa MA, Ghersa CM (2011) Neotyphodium endophyte transmission to Lolium multiflorum seeds depends on the host plant fitness. Environmental and Experimental Botany 71: 359-366. doi https://doi.org/10.1016/j.envexpbot.2011.02.002 (Note to the editor, "Neotyphodium" is a now disfavored name for asexual species of Epichloë.) It might be interesting to look into whether the data on proportions of positive seeds in positive samples suggest stress-induced reductions in vertical transmission, though that would probably be something for future research. But I think that the potential for such an effect should receive some discussion here.}
\response{We agree that evaluating potential effects of stress on imperfect transmission would be very interesting! We feel, based on previous work from our lab group and others (Sneck et al., {Microbial Ecology}, 2017; Cavazos et al., \emph{New Phytologist}, 2018; Gagic et al., \emph{Frontiers in Plant Science}, 2018), that doing so properly would require evaluating larger numbers of seeds per individual to quantify with greater precision the rate of imperfect transmission. 
We scored up to five seeds per specimen, which was adequate for assessing symbiont status but probably too low to provide robust estimates for the seed transmission rate of symbiotic specimens. 
Therefore, we have opted to incorporate this material within the discussion, suggesting future targeted sampling that could be done to get these data from historic specimens.}
\linesref{ms-R5C5-begin}{ms-R5C5-end}

\comment{I also didn't see mention of the potential for horizontal transmission. Some \emph{Epichloë} species on some host "choke" the hosts, meaning that they produce their fruiting structures (stromata) and consequently can spread via airborne ascospores. The three associations investigated here (two \emph{Agrostis} spp. probably with \emph{Epichloe amarillans} and \emph{Elymus virginicus} probably with \emph{Epichloe elymi}) all exhibit this capability even though seed transmission does appear to be the more common route for them.}
\response{Reviewer 5 is correct that there is a possibility of horizontal transmission in general for \emph{Epichloë}, and indeed for these host species. In response to this comment, we acknowledge the potential for horizontal transmission in our description of the study system, and where relevant within the the discussion section. In particular, we note that stromata production has been rare when it has been observed for these host species (less than 1\% of individuals), and also we acknowledge how horizontal transmission would or would not influence our interpretation -- If horizontal transmission were very common, the rising symbiont prevalence we observed in our study could be a signature of symbionts becoming more parasitic and dispersing more commonly through infectious spread, rather than a strengthening of mutualism.}
\linesref{ms-R5C3-begin}{ms-R5C3-end}
\linesref{ms-R5C5-b-begin}{ms-R5C5-b-end}
\linesref{ms-R5C5-discussion-begin}{ms-R5C5-discussion-end}

### 2025-10-06 Manuscript/endo_herbarium_R3.tex
committing manuscript and response updates

\newcommand{\firstrevise}[1]{{\color{black}{#1}}}
\newcommand{\secondrevise}[1]{{\color{black}{#1}}}
\newcommand{\revise}[1]{{\color{Mahogany}{#1}}}

@@

\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, \linelabel{R5C7-begin-dubious}\revise{estimated from surveys to associate with $\sim 20 - 30$\%} of species \revise{in the diverse Poaceae family} \citep{leuchtmann1992systematics}. \revise{Surveys within the cool-season grass subfamily (Pooideae) have found that associations occur for between ~17\% to ~40\% of sampled species \cite{iannone2011diversity,card2014mutualistic}}. \linelabel{R5C7-end-dubious}
They are \firstrevise{predominantly} transmitted vertically from maternal plants to offspring through seeds.
Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
Over time, endophytes that act as mutualists should rise in prevalence within a host population, particularly under environmental conditions that elicit protective benefits \citep{donald2021context}. 
\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic} and protect their hosts against herbivores \linelabel{R5C8-discussion-begin}\revise{\cite{crawford2010fungal,ambrose2014horizontal}} and pathogens \revise{\cite{xia2018role,tian2017epichloe}} likely through the production of a suite of  \linelabel{R5-C9a-begin}\revise{biologically active molecules, including diverse} alkaloids, \revise{proteins,} and other secondary metabolites.\linelabel{R5-C9a-end}\linelabel{R5C8-discussion-end}
The fitness feedback induced by vertical transmission leads to the prediction that endophyte prevalence should be high in populations where these fitness benefits are most important. 
Previous survey studies of contemporary populations have documented large-scale spatial patterns in endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
We predicted that endophyte prevalence should track temporal changes in environmental drivers \linelabel{R2C21-begin}\firstrevise{(i.e. drought)} \linelabel{R2C21-end} that elicit strong fitness benefits.
%For example, endophyte-mediated drought tolerance should lead endophyte prevalence to increase in regions where precipitation has declined over time.

@@

\linelabel{R1C3-begin}
\firstrevise{Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, and \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
The fungal symbionts primarily reproduce asexually and are passed from maternal plant to offspring by vertical transmission through seeds. 
\linelabel{R5C3-begin}\revise{These symbionts are also capable of horizontal transmission between hosts via the production of external reproductive structures, including sexual spore-bearing stromata that grow over host inflorescences (known as 'choke disease') and epiphyllous conidia on leaf surfaces that produce asexual spores \cite{tadych2014epichloe}. 
Evidence suggests that production of horizontal transmission structures by \emph{Epichloë} occurs at low levels, and may be influenced by environmental and genotypic factors \cite{brem1999high,meijer2000effects,tintjer2008variation}.
In line with this, monitoring of long-term plots of \emph{A. perennans} and \emph{E. virginicus} showed no production of stromata on \emph{A. perennans} and on only less than 1\% of \emph{E. virginicus} plants over 14 years \cite{fowler2024microbial}}.\linelabel{R5C3-end}
Some host species have shown \revise{the capacity} to partner with multiple symbiont species or strains, and in some cases multiple symbiont lineages can co-exist within a host population \cite{mc2014species}.
However, surveys have typically found limited \emph{Epichloë} genotypic diversity within host populations \cite{treindl2023genetic}. 
Across host populations, concentrations of biologically-active \revise{biomolecules} and the genes associated with their production vary substantially \cite{schardl2012chemotypic}.
In this analysis, we focus on the presence/absence of \emph{Epichloë} symbionts, and we discuss potential implications of symbiont genotypic diversity in the Discussion.}
\linelabel{R1C3-end}

@@

\linelabel{R2C40-begin}Our examination of historic plant specimens revealed \firstrevise{previously hidden} \linelabel{R2C40-end}shifts in microbial symbiosis over the last two centuries. 
For the three grass host species we examined, there have been strong increases in prevalence of\linelabel{R2C41-begin} \firstrevise{\emph{Epichloë} endophyte symbiosis}.\linelabel{R2C41-end}
We interpret increases in prevalence of \emph{Epichloë}, which are vertically transmitted, as adaptive changes that improve the fitness of their hosts under increasing environmental stress.
This interpretation is in line with theory predicting that positive fitness feedback caused by vertical transmission leads beneficial symbionts to rise in prevalence within a population \cite{fine1975vectors,donald2021context}.
We further found that trends in endophyte prevalence often varied across the host distribution in association with changes in climate drivers, consistent with the hypothesis that increases in endophyte prevalence are driven by context-dependent benefits to hosts that confer resilience under environmental change.
Taken together, our results suggest an overall strengthening of host-symbiont mutualism over the last two centuries.

@@

\firstrevise{While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.}
\linelabel{R2C45-begin}First, climate change responses may depend on genotype-specific responses that are not considered in our current analysis. 
While \emph{Epichloë} symbioses are highly specialized, surveys have demonstrated genotypic and chemotypic diversity of the symbionts among and within populations \cite{von2021genetic, treindl2023genetic}.
Genotypic variation in \emph{Epichloë} endophytes, particularly in genes responsible for alkaloid production, produces ``chemotypes" with differing benefits for hosts against insect or mammalian herbivores mediated by environmental conditions \cite{ambrose2014horizontal,schardl2012chemotypic,saikkonen2013chemical}.  
Genotypic variation of the hosts themselves can also influence interaction outcomes \cite{gundel2011interaction,parker2017genotype}.
Whether hotspots of change in endophyte prevalence reflect selection for genotype-pairings with particularly strong fitness benefits is an unanswered question.\linelabel{R2C45-end}
\linelabel{R2C44-begin}Additionally, \emph{Epichloë} endophytes have been connected to a suite of non-drought related fitness benefits including herbivory \firstrevise{defense} \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of pathogens \cite{vikuk2019infection} and the soil microbiome \citep{roberts2015rhizosphere}. 
\firstrevise{Broad changes in the distribution and abundance of natural enemies \cite{cote2004ecological}, along with stresses from anthropogenic changes in \linelabel{R3C7-begin}\secondrevise{land management} and pollution \cite{sage2020global} likely influence the benefits of symbiosis \cite{rudgers2020climate}.}
\secondrevise{Changing endophyte prevalence} results from the combination of \secondrevise{net} fitness benefits playing out across the heterogeneous map of a changing climate and \linelabel{R3C8-begin}\secondrevise{and its interactive effects on} other anthropogenic drivers.\linelabel{R2C44-end}\linelabel{R3C8-end}
\secondrevise{Host species experience a world that is made increasingly stressful by a combination of global change drivers, and while historic trends that we observed suggest that symbiotic fitness benefits have helped mitigate this stress, it is possible that at yet higher levels of stress, increasing costs of the mutualism could lead to declines in endophyte prevalence. 
\linelabel{R5C5-begin}\revise{It is also know that stressful conditions can reduce the rate of successful transmission of \emph{Epichloë} endophytes from mother plant to offspring \cite{gundel2011neotyphodium}, that may result from host sanctions on costly symbionts or symbiont failure to successfully colonize seeds \cite{afkhami2008symbiosis,gundel2011incorporating}. 
While we did not investigate differences in the rate of transmission across these historic specimens due to low numbers of sampled seeds per individual, it would be extremely valuable to conduct deeper sampling for a subset of specimens with known endophyte status and investigate how transmission itself may respond to environmental change.}\linelabel{R5C5-end}
More extreme climate stresses, which are expected more frequently in the future \cite{seneviratne2021weather} could shift the balance of interactions costs and benefits.
Identifying `tipping points' of mutualism breakdown under increasing environmental stress is an important area of future inquiry.}\linelabel{R3C7-end}

Our results indicate that \emph{Epichloë} symbiosis has likely improved host fitness in stressful environments leading to increasing prevalence. 
\linelabel{R2C42-begin}What is less clear is how this will influence future range shifts.  
Based on our analysis, it is likely that the symbiosis will facilitate range shifts for hosts by improving population growth at range edges.
Previous population surveys \citep{sneck2017variation,rudgers2009benefits,semmartin2015broad} attributed environment-dependent gradients in endophyte prevalence to symbiont-derived fitness benefit's allowing hosts to persist in environments where they otherwise could not \citep{afkhami2014mutualist,kazenel2015mutualistic}.
\linelabel{R1C11-b-begin}However, symbiont-facilitated range shifts require that endophytes be present in the populations to be able to \secondrevise{contribute to} population growth. 
\linelabel{R5C7-begin}For example, the arid western range edge of \emph{A. hyemalis} has had historically low endophyte prevalence (Fig. \ref{fig:svc_space_year_map})\revise{, and dispersal of symbionts may limit the capacity for range shifts. }
\revise{A range edge population with no endophytes would require dispersal of symbionts to occur, either via symbiotic seeds or \linelabel{R5C5-discussion-begin} horizontally-transmitted fungal spores, before the symbiosis could result in an expanding distibution \cite{fowler2023geographic}}\linelabel{R5C5-discussion-end} 
\revise{At the same time, we found that} endophyte prevalence has increased most quickly in regions with historically low endophyte prevalence (Fig. \ref{fig:initialprev_trend_plot}), suggesting strong selection for symbiotic hosts when they are present.\linelabel{R1C11-b-end}\linelabel{R5C7-end}
These factors potentially contribute to the ability of the host species to track its environmental niche. 
Another interesting question is the degree to which symbiotic and non-symbiotic hosts, which occupy overlapping but distinct niches, are likely to experience distribution shifts in tandem or at different rates in the future.\linelabel{R2C42-end}

>>>>

\newcommand{\firstrevise}[1]{{\color{black}{#1}}}
\newcommand{\secondrevise}[1]{{\color{black}{#1}}}
\newcommand{\revise}[1]{{\color{black}{#1}}}

@@

\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, \linelabel{R5C7-begin-dubious}\revise{estimated from surveys to associate with $\sim 20 - 30$\%} of species \revise{across the diverse Poaceae family} \citep{leuchtmann1992systematics}. 
\revise{Within the cool-season grass subfamily (Pooideae), it has been estimated that between 17\% to 40\% of sampled species act as \emph{Epichloë} hosts \cite{iannone2011diversity,card2014mutualistic}}. \linelabel{R5C7-end-dubious}
They are \firstrevise{predominantly} transmitted vertically from maternal plants to offspring through seeds.
Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
Over time, endophytes that act as mutualists should rise in prevalence within a host population, particularly under environmental conditions that elicit protective benefits \citep{donald2021context}. 
\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic} and protect their hosts against herbivores \linelabel{R5C8-begin}\revise{\cite{crawford2010fungal,ambrose2014horizontal}} and pathogens \revise{\cite{xia2018role,tian2017epichloe}} likely through the production of a suite of  \linelabel{R5-C9a-begin}\revise{biologically active molecules, including diverse} alkaloids, \revise{proteins,} and other secondary metabolites.\linelabel{R5-C9a-end}\linelabel{R5C8-end}
The fitness feedback induced by vertical transmission leads to the prediction that endophyte prevalence should be high in populations where these fitness benefits are most important. 
Previous survey studies of contemporary populations have documented large-scale spatial patterns in endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
We predicted that endophyte prevalence should track temporal changes in environmental drivers \linelabel{R2C21-begin}\firstrevise{(i.e. drought)} \linelabel{R2C21-end} that elicit strong fitness benefits.
%For example, endophyte-mediated drought tolerance should lead endophyte prevalence to increase in regions where precipitation has declined over time.

@@

\linelabel{R1C3-begin}
\firstrevise{Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, and \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
The fungal symbionts primarily reproduce asexually and are passed from maternal plant to offspring by vertical transmission through seeds. 
\linelabel{R5C3-begin}\revise{These symbionts are also capable of horizontal transmission between hosts via the production of external reproductive structures, including sexual spore-bearing stromata that grow over host inflorescences (known as 'choke disease') and epiphyllous conidia on leaf surfaces that produce asexual spores \cite{tadych2014epichloe}. 
Evidence suggests that production of horizontal transmission structures by \emph{Epichloë} occurs at low levels, and may be influenced by environmental and genotypic factors \cite{brem1999high,meijer2000effects,tintjer2008variation}.
In line with this, monitoring of long-term plots of \emph{A. perennans} and \emph{E. virginicus} showed no production of stromata on \emph{A. perennans} and on only less than 1\% of \emph{E. virginicus} plants over 14 years \cite{fowler2024microbial}}.
A similar low frequency of stromata formation (only 0.37\% of recorded inflorescences) was observed for \emph{A. hyemalis} in a separate field experiment \cite{donald2021context}.\linelabel{R5C3-end}
Some host species have shown \revise{the capacity} to partner with multiple symbiont species or strains, and in some cases multiple symbiont lineages can co-exist within a host population \cite{mc2014species}.
However, surveys have typically found limited \emph{Epichloë} genotypic diversity within host populations \cite{treindl2023genetic}. 
Across host populations, concentrations of biologically-active \revise{biomolecules} and the genes associated with their production vary substantially \cite{schardl2012chemotypic}.
In this analysis, we focus on the presence/absence of \emph{Epichloë} symbionts, and we discuss potential implications of symbiont genotypic diversity in the Discussion.}
\linelabel{R1C3-end}

@@

\linelabel{R2C40-begin}Our examination of historic plant specimens revealed \firstrevise{previously hidden} \linelabel{R2C40-end}shifts in microbial symbiosis over the last two centuries. 
For the three grass host species we examined, there have been strong increases in prevalence of\linelabel{R2C41-begin} \firstrevise{\emph{Epichloë} endophyte symbiosis}.\linelabel{R2C41-end}
We interpret increases in prevalence of \emph{Epichloë}, which are \revise{predominantly} vertically transmitted, as adaptive changes that improve the fitness of their hosts under increasing environmental stress.
This interpretation is in line with theory predicting that positive fitness feedback caused by vertical transmission leads beneficial symbionts to rise in prevalence within a population \cite{fine1975vectors,donald2021context}.
We further found that trends in endophyte prevalence often varied across the host distribution in association with changes in climate drivers, consistent with the hypothesis that increases in endophyte prevalence are driven by context-dependent benefits to hosts that confer resilience under environmental change.
Taken together, our results suggest an overall strengthening of host-symbiont mutualism over the last two centuries.

@@

\firstrevise{While we show consistent increasing trends in prevalence between the three species, the mechanisms that explain these changes may be diverse and idiosyncratic.}
\linelabel{R2C45-begin}First, climate change responses may depend on genotype-specific responses that are not considered in our current analysis. 
While \emph{Epichloë} symbioses are highly specialized, surveys have demonstrated genotypic and chemotypic diversity of the symbionts among and within populations \cite{von2021genetic, treindl2023genetic}.
Genotypic variation in \emph{Epichloë} endophytes, particularly in genes responsible for alkaloid production, produces ``chemotypes" with differing benefits for hosts against insect or mammalian herbivores mediated by environmental conditions \cite{ambrose2014horizontal,schardl2012chemotypic,saikkonen2013chemical}.  
Genotypic variation of the hosts themselves can also influence interaction outcomes \cite{gundel2011interaction,parker2017genotype}.
Whether hotspots of change in endophyte prevalence reflect selection for genotype-pairings with particularly strong fitness benefits is an unanswered question.\linelabel{R2C45-end}
\linelabel{R2C44-begin}Additionally, \emph{Epichloë} endophytes have been connected to a suite of non-drought related fitness benefits including herbivory \firstrevise{defense} \citep{brem2001epichloe}, salinity resistence \citep{wang2020effects}, and mediation of pathogens \cite{vikuk2019infection} and the soil microbiome \citep{roberts2015rhizosphere}. 
\firstrevise{Broad changes in the distribution and abundance of natural enemies \cite{cote2004ecological}, along with stresses from anthropogenic changes in \linelabel{R3C7-begin}\secondrevise{land management} and pollution \cite{sage2020global} likely influence the benefits of symbiosis \cite{rudgers2020climate}.}
\secondrevise{Changing endophyte prevalence} results from the combination of \secondrevise{net} fitness benefits playing out across the heterogeneous map of a changing climate and \linelabel{R3C8-begin}\secondrevise{and its interactive effects on} other anthropogenic drivers.\linelabel{R2C44-end}\linelabel{R3C8-end}
\secondrevise{Host species experience a world that is made increasingly stressful by a combination of global change drivers, and while historic trends that we observed suggest that symbiotic fitness benefits have helped mitigate this stress, it is possible that at yet higher levels of stress, increasing costs of the mutualism could lead to declines in endophyte prevalence. 
\linelabel{R5C5-begin}\revise{It is also known that stressful conditions can both reduce \cite{gundel2011neotyphodium} and increase \cite{gundel2020simulated} the rate of successful transmission of \emph{Epichloë} endophytes from mother plant to offspring.
These responses likely reflect both consequences of host sanctions on costly symbionts and the ability of symbionts to successfully colonize seeds \cite{afkhami2008symbiosis,gundel2011incorporating}. 
While we did not investigate differences in the rate of transmission across these historic specimens due to low numbers of sampled seeds per individual, it would be valuable to conduct deeper sampling for a subset of specimens with known endophyte status and investigate how transmission itself may respond to environmental change.}\linelabel{R5C5-end}
\linelabel{R5C5-b-begin}\revise{Additionally, we have interpreted increasing prevalence as a signature of increased host fitness driven by vertical transmission, which we expect is the predominant mode of tranmission in this system. 
Horizontal transmission, while rare, could also explain increasing prevalence if global change stressors lead symbionts to engage in increased parasitic behavior, resulting in increased disperal and transmission at the cost of host fitness.}\linelabel{R5C5-b-end}
More extreme climate stresses, which are expected more frequently in the future \cite{seneviratne2021weather} could shift the balance of interactions costs and benefits.
Identifying `tipping points' of mutualism breakdown under increasing environmental stress is an important area of future inquiry.}\linelabel{R3C7-end}

Our results indicate that \emph{Epichloë} symbiosis has likely improved host fitness in stressful environments leading to increasing prevalence. 
\linelabel{R2C42-begin}What is less clear is how this will influence future range shifts.  
Based on our analysis, it is likely that the symbiosis will facilitate range shifts for hosts by improving population growth at range edges.
Previous population surveys \citep{sneck2017variation,rudgers2009benefits,semmartin2015broad} attributed environment-dependent gradients in endophyte prevalence to symbiont-derived fitness benefit's allowing hosts to persist in environments where they otherwise could not \citep{afkhami2014mutualist,kazenel2015mutualistic}.
\linelabel{R1C11-b-begin}However, symbiont-facilitated range shifts require that endophytes be present in the populations to be able to \secondrevise{contribute to} population growth. 
\linelabel{R5C7-begin}For example, the arid western range edge of \emph{A. hyemalis} has had historically low endophyte prevalence (Fig. \ref{fig:svc_space_year_map})\revise{, and dispersal of symbionts may limit the capacity for range shifts. }
\revise{A range edge population with no endophytes would require dispersal of symbionts to occur, either via symbiotic seeds or \linelabel{R5C5-discussion-begin} horizontally-transmitted fungal spores, before the symbiosis could result in an expanding distribution \cite{fowler2023geographic}.}\linelabel{R5C5-discussion-end} 
\revise{At the same time, we found that} endophyte prevalence has increased most quickly in regions with historically low endophyte prevalence (Fig. \ref{fig:initialprev_trend_plot}), suggesting strong selection for symbiotic hosts when they are present.\linelabel{R1C11-b-end}\linelabel{R5C7-end}
These factors potentially contribute to the ability of the host species to track its environmental niche. 
Another interesting question is the degree to which symbiotic and non-symbiotic hosts, which occupy overlapping but distinct niches, are likely to experience distribution shifts in tandem or at different rates in the future.\linelabel{R2C42-end}
